Combination of field and experimental data together with computational models reveal cognitive mechanisms behind cleaner fish behaviour

### 2022-01-14 manuscript_1.0/manuscript_1.0.tex
New Rmd file to produce the manuscript in word format. Abstract and intro have been edited as well as figures. In the process of editing methods and discussion.

%%%% Abstract text to be placed here %%%%%%%%%%%%
\begin{abstract}
The same behaviour can be achieved by different mechanistic underpinnings. But the environmental conditions under which the behaviour is triggered might differ between these potential mechanisms. Thus, by analyzing the varitation of expression of a certain behavior across environmental gradients, it is possible to infer which mechanism is behind the behaviour of interest. To do this it is necessary to have clearcut predictions for the different candidate mechanisms. In here, we use experimental data on the behaviour of the cleaner fish \emph{Labroides} \emph{dimidiatus}, as well as enviromental variables from the location of fish, together with a computational model of the learning mechanisms. Using bayesian statistics, we fit key parameters of the computational model that determine the specific learning mechanism. We find that a model where cleaners estimate the long-term consequences of their actions, and use the estimate in their desicion making, explains better under which conditions the fish develop a preference for an ephemeral food source in the context of a biological market. We argue that the combination of computational models with data is powerful tool to infer the mechanistic underpinning behind behaviours of interest.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Cooperation among individuals of the same or different species is
ubiquitous in nature. Among individuals of different species it
underlays the biological mechanism for fundamental ecological
interactions such as pollination, nitrogen fixation or parasite removal
\citep{begon_Ecology_2006}. When partners in these mutualistic interactions
are able to choose their interaction partner, their investment and the
return they obtain can be understood as a biological market
\citep{hammerstein_Biological_2016, leimar_Cooperation_2010, noe_Biological_1994a, noe_Biological_1995a}. Thus, we expect supply
and demand of the good and services being exchanged to determine the
action of individuals, at least in an evolutionary time-scale. However,
supply and demand conditions of a biological market, which typically
depend on the abundance of the species involved, certainly vary in time
and space. How do partners adjust their strategy in the face of this
variation? In animals an obvious general candidate mechanism for the
strategic adjustment is the cognitive machinery. However, it is not
clear which specific cognitive mechanisms allow individual to adjust
their behaviour to the varying market conditions.

One example of strategic adjustment in a biological market is the marine
cleaning mutualism involving the cleaner fish \emph{Labroides dimidiatus} and
`client' fish. Clients offer themselves as food patches for cleaners to
remove ectoparasites, which gives the cleaners calories and the clients
improved health \citep{waldie_LongTerm_2011, ros_Does_2011, triki_Effects_2016}. As cleaners have about 2000
interactions per day \citep{grutter_Relationship_1995}, they regularly face
the situation that two clients seek cleaning simultaneously. In such
conditions, it is important to distinguish between `resident' and
`visitor' client species \citep{bshary_Cleaner_2001a}. The residents have
small home ranges and hence access to only the local cleaner, while the
visitors have large home ranges and hence access to two or more
cleaners. Thus, visitors have the option to switch to another cleaner if
being made to wait, while residents must wait for inspection. Indeed,
visitors have been observed to use their partner choice options in that
way \citep{bshary_Choosy_2002}, which may explain why cleaners give
visitors service priority in a field study in the Red Sea
\citep{bshary_Cleaner_2001a}. More generally, cleaners apparently prefer to
interact with visitors \citep{triki_Marine_2021}. Such a preference may be
linked to a client's value as a food patch; visitors are on average
larger than residents \citep{bshary_Cleaner_2001a}, and size is positively
correlated with parasite load \citep{grutter_Relationship_1995a}. However,
early laboratory
experiments showed that leaving causes service priority independent of
client value \citep{bshary_Asymmetric_2002}. The experiments consisted of a
simultaneous two-choice task, where cleaners had to decide which of two
Plexiglas plates, which offered equal amounts of food, to approach first
and eat mashed prawn of it. One plate was made to behave like a
resident, i.e.~it would remain until the cleaner had eaten off it. The
other plate was made to behave like a visitor, i.e.~it would be removed
if not inspected first. Cleaners learned to prefer the visitor plate and
hence obtained the double amount of food \citep{bshary_Asymmetric_2002}.
Adult cleaner fish outperform various primates as well as rats and
pigeons in this original version of the biological market task
\citep{salwiczek_Adult_2012, zentall_Early_2017}; only African grey
parrots solve this task as well \citep{pepperberg_Can_2014}.

Recent research has refined our view on partner choice and priority of
access in the \emph{L. dimidiatus} cleaning mutualism, and as a consequence
performance in the biological market task. First, details of the task
presentation may matter; i.e.~what relevant information is encoded in
the design of the two plates (colour versus size;
\citep{wismer_Cuebased_2019}), or whether the information is in the colour of
the food \citep{pretot_Comparative_2021, pretot_Comparing_2016}, and
whether the food is visible or hidden/released only when the subjects
made choices on a touch screen \citep{pretot_Factors_2016}. Second, it
appears that cleaner and client densities in nature affect how cleaners
perform in the original market task: when densities are low, cleaners
are more likely to fail to complete the task \citep{triki_Biological_2019, triki_Decrease_2018, wismer_Variation_2014}. The apparent explanation
for this failure is that visitor clients are less likely to exert their
choice options (and hence wait for inspection if made to wait) when
cleaner fish densities are low \citep{triki_Brain_2020}. This change in
behaviour may be due to the fact that when cleaner densities are low,
the cost of switching in terms of distance to swim and time invested
increases. If visitors are not choosy, cleaners should not give them
priority anymore. Interestingly, while cleaners from high-density areas
have larger forebrains harbouring more cells than cleaners from
low-density areas \citep{triki_Brain_2020}, on a local scale those
individuals that perform in the market task in a way that corresponds to
the locally optimal strategy (solving/failing the task when living in
high/low density) have relatively larger forebrains \citep{triki_Brain_2020}.
Taken together, empirical results clearly show major intra- and
inter-specific variation in performance in the market task, warranting
investigation of the underlying cognitive machinery, from perception to
decision-making \citep{shettleworth_Cognition_2009}.

One approach to identify candidate mechanisms underlying performance in
a cognitive task is to develop proximate learning models and to evaluate
their predictions in light of the behavioural data (e.g.
\citep{dubois_Model_2021}). By varying the cognitive tool kit, one can assess
which minimal kit is necessary to solve the task at hand. This approach
has been used to analyse the market task \citep{prat_Role_2021, quinones_Reinforcement_2019}. Learning models showed that basic
reinforcement learning does not suffice to solve the task, this is
particularly so when models assume the more complex natural situation in
which cleaners not only face resident-visitor pairs but also
visitor-visitor and resident-resident pairs, as well only a resident or
only a visitor. To be able to give visitors priority over residents,
cleaners first need to be able to assess a client's value separately for
the three possible forms of encounter (alone, paired with a fish with
the same strategic option, paired with a fish with the alternative
strategic option) \citep{quinones_Reinforcement_2019}. In addition, cleaners
need to account for the future consequences of current decisions. In the
model, this could be achieved in two non-mutually exclusive ways
\citep{quinones_Reinforcement_2019}: through low temporal discounting of
future effects and/or through perceiving a visitor client leaving as
psychological punishment (i.e.~as a negative reinforcer). Given that a
negative reinforcer is part of the ubiquitous learning mechanism termed
operant conditioning \citep{thorndike_Animal_1898, skinner_Behavior_1938},
this explanation can be
considered the simpler one. Furthermore, accounting for the future
effects of an action is considered an advanced cognitive ability
\citep{suddendorf_Evolution_2007}. Here, we fitted our model to the empirical
data on fish densities and cleaner fish performance in the market task
with the aim to determine the likely mechanism used by cleaners to
incorporate future consequences of current decisions.

>>>>

%%%% Abstract text to be placed here %%%%%%%%%%%%
\begin{abstract}
While it is generally straightforward to quantify individual performance in cognitive experiments, identifying the underlying cognitive processes remains a major challenge. Often, different mechanistic underpinnings may yield similar performances, and Lloyd Morgan's cannon warrants acceptance of the simpler explanation. Alternatively, when the different mechanisms interact with environmental conditions, variation in performance across environments might allow to statistically infer the mechanism responsible. However, to do this it is necessary to have quantitative predictions for the different candidate mechanisms. Here, we use a set of computational models to get quantitative predictions from alternative learning mechanisms, and using Bayesian statistics fit the model parameters to performance data. We used experimental data on performance in an ephemeral reward task by wild-caught cleaner fish Labroides dimidiatus, as well as cleaner and client fish densities from the locations of capture. The task can in principle be solved by estimating future consequences of an action, or by perceiving the removal of the ephemeral reward as psychological punishment (negative reinforcement). We found that a model where cleaners only estimate the future consequences of their actions explains best which cleaner relative abundances cause the fish to develop a preference for an ephemeral food source. This model also yields performances that can be considered the result of locally optimal decision-rules, in contrast to the negative reinforcement model. We argue that the combination of computational models with data is a powerful tool to infer the mechanistic underpinning behind cognitive performance.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Cooperation among individuals of the same or different species is
ubiquitous in nature. Whenever such interactions involve some degree
of partner choice, biological market theory predicts that the exchange
rate of goods and/or services traded adjusts to the law of supply and demand
\citep{noe_Biological_1995a}. While biological market models have largely focused
on evolutionary time scales \citep{noe_Biological_1994a, johnstone_Mutualism_2008a}
\(\color{red}{\text{Hoekstra xxxx}}\), supply and demand conditions, which typically
depend on the abundance of the species involved, certainly vary in time
and space within ecological time scales. Therefore, natural selection should
favour the ability to flexibly adjust decisions and behavioural output to
current market conditions. Indeed, such adjustments have been repeatedly
documented \citep{axen_Signalling_1996}. In animals, an obvious general candidate
mechanism for the strategic adjustment is the cognitive machinery. However,
it is not clear which cognitive mechanisms allow individuals to adjust their
behaviour to the varying market conditions. As adjustments to changes in
market conditions have been documented in a great variety of species with
highly variable brain anatomies, the question arises to what extent mechanisms
beyond basic associative learning may be involved.

One example of strategic adjustment in a biological market is the marine
cleaning mutualism involving the cleaner fish \emph{Labroides dimidiatus} and
`client' fish. Clients offer themselves as food patches for cleaners to
remove ectoparasites, which gives the cleaners calories and the clients
improved health \citep{waldie_LongTerm_2011, ros_Does_2011, triki_Effects_2016}. As cleaners have about 2000
interactions per day \citep{grutter_Relationship_1995}, they regularly face
the situation that two clients seek cleaning simultaneously. In such
conditions, it is important to distinguish between `resident' and
`visitor' client species \citep{bshary_Cleaner_2001a}. The residents have
small home ranges and hence access to only the local cleaner, while the
visitors have large home ranges and hence access to two or more
cleaners. Thus, visitors have the option to switch to another cleaner if
being made to wait, while residents must wait for inspection. Indeed,
visitors have been observed to use their partner choice options in that
way \citep{bshary_Choosy_2002}, which may explain why cleaners give
visitors service priority in a field study in the Red Sea
\citep{bshary_Cleaner_2001a}. More generally, cleaners apparently prefer to
interact with visitors \citep{triki_Marine_2021}. Such a preference may be
linked to a client's value as a food patch; visitors are on average
larger than residents \citep{bshary_Cleaner_2001a}, and size is positively
correlated with parasite load \citep{grutter_Relationship_1995a}. However,
early laboratory experiments showed that leaving causes service
priority independent of client value \citep{bshary_Asymmetric_2002}.
The experiments consisted of asimultaneous two-choice task,
where cleaners had to decide which of two
Plexiglas plates, which offered equal amounts of food, to approach first
and eat mashed prawn of it. One plate was made to behave like a
resident, i.e.~it would remain until the cleaner had eaten off it. The
other plate was made to behave like a visitor, i.e.~it would be removed
if not inspected first. Cleaners learned to prefer the visitor plate and
hence obtained the double amount of food \citep{bshary_Asymmetric_2002}.
Adult cleaner fish outperform various primates as well as rats and
pigeons in this original version of the biological market task
\citep{salwiczek_Adult_2012, zentall_Early_2017}; only African grey
parrots solve this task as well \citep{pepperberg_Can_2014}. As the correct
solution must be learned, and primates are known to readily distinguish
between one and two reward units if presented in more conventional ways
\(\color{red}{\text{ref}}\), what makes the task difficult for most non-cleaners?

Recent research on the cognitive tool kit needed to solve an ephemeral
reward task has used a broad approach as proposed by Shettleworth
\citep{shettleworth_Cognition_2009},
who defined cognition as includes all ways in which animals take in information
through the senses, process, retain and decide to act on it. Indeed, perception
of relevant cues is of major importance for individual performance;
i.e.~what relevant information is encoded in the design of the two
plates (colour versus size; \citep{wismer_Cuebased_2019}), or whether
the information is in the colour of
the food \citep{pretot_Comparative_2021, pretot_Comparing_2016}, and
whether the food is visible or hidden/released only when the subjects
made choices on a touch screen \citep{pretot_Factors_2016}. However, identifying
cues as salient is not the only challenge of the ephemeral reward task,
as revealed by proximate learning models. Such models allow varying the
cognitive tool kit and to evaluate which minimal kit is necessary to solve
the task at hand (e.g. \citep{dubois_Model_2021}). Applied to the ephemeral
reward task, learning models showed that basic reinforcement learning does
not suffice to solve the task \citep{prat_Modelling_2022, quinones_Reinforcement_2019}.
This is particularly so when models assume the more complex natural
situation in which cleaners not only face resident-visitor pairs but also
visitor-visitor and resident-resident pairs, as well only a resident or
only a visitor. To be able to give visitors priority over residents,
cleaners need to be able to assess a client's value separately for the
three possible forms of encounter (alone, paired with a fish with the
same strategic option, paired with a fish with the alternative strategic option)
\citep{quinones_Reinforcement_2019}. The ability to keep such separate scores
for the same stimulus has been termed configurational learning,
or chunking, or segmentation (see references in \citep{prat_Modelling_2022}).

To solve the ephemeral reward task, cleaners also need to account for
the future consequences of current decisions. In the model by
Quiñones et al. \citep{quinones_Reinforcement_2019}, this could be achieved in
two non-mutually exclusive ways: through low temporal discounting of
future effects, also termed `chaining' \citep{enquist_Power_2016}; and/or
through perceiving a visitor client leaving as psychological punishment
(i.e.~as a negative reinforcer). Given that a negative reinforcer is part
of the ubiquitous learning mechanism termed operant conditioning
\citep{thorndike_Animal_1898, skinner_Behavior_1938}, this explanation
could be considered the simpler one. Furthermore, accounting for the
future effects of an action is considered an advanced cognitive
ability \citep{suddendorf_Evolution_2007}. Thus, standard logic of
Lloyd Morgan's cannon demands that operant conditioning as the simpler
explanation is to be accepted by default. Here, we explore the
possibility that fitting a model in which the action of these two
mechanisms, operant conditioning and chaining, depends
on model parameters may reveal which process cleaners actually use.
This is because one process may be better at predicting the observed
variation in cleaner fish performance as a function of ecological conditions.

Various repetitions of the ephemeral reward task in different years and
with cleaners caught from various locations around Lizard Island,
Great Barrier Reef, revealed that cleaner and client densities in
nature affect how cleaners perform in the original market task:
when relative densities are low, cleaners are more likely to fail to
complete the task \citep{triki_Biological_2019, triki_Decrease_2018, wismer_Variation_2014}. The explanation for this pattern
is twofold. First, low cleaner abundances implies low supply and high demand
of cleaning services. Under this conditions, there will be less occasions under
which failing to give priority to a visitor translates to an empty
cleaning station. Secondly and also related to the state of the cleaning market,
visitor clients are less likely to exert their choice options
(and hence wait for inspection if made to wait) when
cleaner fish densities are low \citep{triki_Brain_2020}. If visitors are not choosy,
cleaners should not give them priority anymore. Thus, the underling
cognitive machinery and ecological variables apparently interact to produce
such variable performance. By fitting our model to the empirical data on
fish densities and cleaner fish performance in the ephemeral reward task
we aimed to address two questions. First, we aimed to determine the likely
mechanism used by cleaners to incorporate future consequences of current
decisions by testing whether operant conditioning, chaining, or a combination
of both best explains performance. Second and related to the first
question, we aimed to determine whether the two mechanisms differed
with respect to the ecological conditions that are likely to cause high
versus low performance in the ephemeral reward task. If that were the case,
we would assess which mechanism would yield more appropriate performance
patterns. Relying on the logic of biological market theory, we predicted
that appropriate performance is to show low preference for visitors under high
local client-to-cleaner ratio as visitor leaving rates should be low
and any leaving visitor will be readily replaced by another client. In contrast,
cleaners should show high performance if the local client-to-cleaner
ratio favours clients.
